Compare label transfer results between Azimuth and Azimuth-adapted strategy

### 2024-11-26 analyses/cell-type-wilms-tumor-06/notebooks/02a_label-transfer_fetal_full_reference_Cao.Rmd
Debugs related to reorg into notebooks, and new data release which can have fewer than 500 cells

---
title: "Reference label transfer of the fetal kidney reference (Stewart et al.) for `r params$sample_id`"
author: "Maud PLASCHKA"
date: '2024-08-13'
params:
  scpca_project_id: "SCPCP000006"
  sample_id: "SCPCS000176"
  seed: 12345
  testing: 0
output: 
  html_document: 
    toc: yes
    toc_float: yes
    code_folding: hide
    highlight: pygments
    df_print: paged
---

@@

In this notebook, we perform label transfer from the human fetal kidney atlas from Stewart et al.

@@

```{r packages, message=FALSE, warning=FALSE}
library(Seurat)
library(sctransform)
library(SCpubr)
library(tidyverse)
library(patchwork)
library(SingleCellExperiment)

@@

```{r base paths, eval=TRUE}
# The base path for the OpenScPCA repository, found by its (hidden) .git directory
repository_base <- rprojroot::find_root(rprojroot::is_git_root)

# The current data directory, found within the repository base directory
data_dir <- file.path(repository_base, "data", "current", params$scpca_project_id)

@@

```{r}
# load functions for label transfer
source(
  file.path(
    module_base,
    "notebook_template",
    "utils",
    "label-transfer-functions.R"
  )
)
```

@@

The reference has been download and pre-process in the `R Script` `download-and-create-fetal-kidney-ref.R`

```{r path_to_reference}
path_to_ref <- file.path(
  module_base,
  "results",
  "references",
  "stewart_formatted_ref.rds"
)
if (!file.exists(path_to_ref)) {
  stop("Reference file could not be found. Make sure `scripts/prepare-fetal-references.R` has been run first.")
}
ref <- readRDS(path_to_ref)

@@

In this notebook, we test the label transfer from the fetal kidney atlas (reference) to the Wilms tumor sample defined in `params$sample_id` from the Wilms tumor dataset `param$scpca_project_id` 
We work with the pre-processed `Seurat`object and labeled with the fetal kidney reference saved in the `results` directory.

@@

Reports will be saved in the `notebook/02-reference` directory. 
The pre-processed and annotated `Seurat` object per samples are saved in the `result` folder.

```{r path_to_output}
output_dir <- file.path(module_base, "results", params$sample_id)
```

@@

### Load the pre-processed `Seurat` object
```{r load, message=FALSE, warning=FALSE}
# open the processed rds object
srat <- readRDS(file.path(data_dir, paste0("02a-fetal_full_label-transfer_", params$sample_id, ".Rds")))
srat_assay <- "RNA"

# prepare the query for label transfer
# we don't want to overwrite the srat object since `prepare_query`
#  removes features that are not present in the reference
DefaultAssay(srat) <- srat_assay
query <- prepare_query(
  srat,
  rownames(reference),
  srat_assay,
  convert_gene_names = FALSE
)
```

### Azimuth annotation from fetal kidney

```{r run_azimuth, message=FALSE, warnings=FALSE}
options(future.globals.maxSize = 8912896000000)

@@

srat <- AddMetaData(srat, metadata_to_transfer, col.name = paste0("fetal_kidney_", annotation_columns))
```

```{r plot_azimuth, fig.height=8, fig.width=8, warnings=FALSE}
d1 <- DimPlot(srat, reduction = "umap", dims = c(1, 2), group.by = "fetal_kidney_predicted.compartment", label = TRUE, repel = TRUE) + NoLegend()
d2 <- DimPlot(srat, reduction = "umap", dims = c(1, 2), group.by = "fetal_kidney_predicted.cell_type", label = TRUE, repel = TRUE) + NoLegend()

f1 <- SCpubr::do_BarPlot(
  sample = srat,
  group.by = "fetal_kidney_predicted.compartment",
  split.by = "seurat_clusters",
  position = "fill",
  font.size = 10,
  legend.ncol = 3
) +
  ggtitle("% cells") +
  xlab(params$sample_id)

f2 <- SCpubr::do_BarPlot(
  sample = srat,
  group.by = "fetal_kidney_predicted.cell_type",
  split.by = "seurat_clusters",
  position = "fill",
  font.size = 10,
  legend.ncol = 3
) +
  ggtitle("% cells") +
  xlab(params$sample_id)

(d1 / f1) | (d2 / f2)
```

Note:

For some reason, the "cap-mesenchyme cells" has been renamed in CELLxGENE as "mesenchymal cells".

The cap mesenchyme is a cap of condensed metanephric mesenchyme, comprised of cells which epithelialize and sequentially form the peritubular aggregate (PA), renal vesicle (RV), C-, and S-shaped bodies, and finally the mature nephron. 
The CM contains nephron progenitor cells.

This can be confusing and we just need to pay attention that :

- the fetal nephron / mesenchymal cells are cap-mesenchyme cells. 
In our case, cap-mesenchyme contains blastema and primitive epithelial cancer cells.

- the stroma / mesenchymal stem are likely mesenchymal cancer or normal cells.

@@

```{r save}
saveRDS(object = srat, file = file.path(output_dir, paste0("02b-fetal_kidney_label-transfer_", params$sample_id, ".Rds")))
```

>>>>

---
title: "Reference label transfer from the fetal full reference (Cao et al.) for `r params$sample_id`"
author: "Maud PLASCHKA"
date: '2024-08-07'
params:
  scpca_project_id: "SCPCP000006"
  sample_id: "SCPCS000176"
  seed: 12345
  homologs_file: "../scratch/homologs.rds"
  testing: 0
output:
  html_document:
    toc: yes
    toc_float: yes
    code_folding: hide
    highlight: pygments
    df_print: paged
---

@@

In this notebook, we perform label transfer from the human fetal Azimuth reference from Cao et al.

@@

```{r packages, message=FALSE, warning=FALSE}
library(Seurat)
library(SeuratData)
library(sctransform)
library(SCpubr)
library(tidyverse)
library(patchwork)
library(SingleCellExperiment)

@@

```{r base paths, eval=TRUE, include=TRUE}
# The base path for the OpenScPCA repository, found by its (hidden) .git directory
repository_base <- rprojroot::find_root(rprojroot::is_git_root)

@@

```{r}
# load functions for label transfer
source(
  file.path(
    module_base,
    "notebooks",
    "utils",
    "label-transfer-functions.R"
  )
)
```

@@

Load the Azimuth reference which has been prepared for label transfer.

```{r path_to_reference}
path_to_ref <- file.path(
  module_base,
  "results",
  "references",
  "cao_formatted_ref.rds"
)
if (!file.exists(path_to_ref)) {
  stop("Reference file could not be found. Make sure `scripts/prepare-fetal-references.R` has been run first.")
}
ref <- readRDS(path_to_ref)

@@

In this notebook, we test the label transfer from the azimuth fetal atlas (reference) to the Wilms tumor sample defined in `params$sample_id` from the Wilms tumor dataset `param$scpca_project_id`.

We work with the pre-processed `Seurat`object saved in the `results` directory.

@@

Reports will be saved in the `notebook/02-reference` directory.
The pre-processed and annotated `Seurat` object per samples are saved in the `result` folder.
```{r path_to_output}
output_dir <- file.path(module_base, "results", params$sample_id)
```

@@

### Load the pre-process data (query)

```{r load, message=FALSE, warning=FALSE}
# open the processed rds object
srat <- readRDS(file.path(data_dir, paste0("01-Seurat_", params$sample_id, ".Rds")))
srat_assay <- "RNA"

# prepare the query for label transfer
# we don't want to overwrite the srat object since `prepare_query`
#  removes features that are not present in the reference
DefaultAssay(srat) <- srat_assay
query <- prepare_query(
  query = srat,
  reference_rownames = rownames(reference),
  assay = srat_assay,
  homolog_file = params$homologs_file
)
```

### Label transfer from fetal kidney reference using Azimuth

```{r run_azimuth, message=FALSE, warnings=FALSE}
options(future.globals.maxSize = 891289600000000)

@@

srat <- AddMetaData(srat, metadata_to_transfer, col.name = paste0("fetal_full_", annotation_columns))

rm(query)
rm(query_labeled)
```

```{r plot_azimuth, fig.height=15, fig.width=8, warnings=FALSE}
d1 <- DimPlot(srat, reduction = "umap", dims = c(1, 2), group.by = "fetal_full_predicted.organ", label = TRUE, repel = TRUE) + NoLegend()
d2 <- DimPlot(srat, reduction = "umap", dims = c(1, 2), group.by = "fetal_full_predicted.annotation.l1", label = TRUE, repel = TRUE) + NoLegend()
d3 <- DimPlot(srat, reduction = "umap", dims = c(1, 2), group.by = "fetal_full_predicted.annotation.l2", label = TRUE, repel = TRUE) + NoLegend()

f1 <- SCpubr::do_BarPlot(
  sample = srat,
  group.by = "fetal_full_predicted.organ",
  split.by = "seurat_clusters",
  position = "fill",
  font.size = 10,
  legend.ncol = 4
) +
  ggtitle("% cells") +
  xlab(params$sample_id)

f2 <- SCpubr::do_BarPlot(
  sample = srat,
  group.by = "fetal_full_predicted.annotation.l1",
  split.by = "seurat_clusters",
  position = "fill",
  font.size = 10,
  legend.ncol = 2
) +
  ggtitle("% cells") +
  xlab(params$sample_id)

f3 <- SCpubr::do_BarPlot(
  sample = srat,
  group.by = "fetal_full_predicted.annotation.l2",
  split.by = "seurat_clusters",
  position = "fill",
  font.size = 10,
  legend.ncol = 2
) +
  ggtitle("% cells") +
  xlab(params$sample_id)

((d1 / f1) | (d2 / f2))
```

@@

```{r save}
saveRDS(object = srat, file = file.path(output_dir, paste0("02a-fetal_full_label-transfer_", params$sample_id, ".Rds")))
```

### 2024-11-26 analyses/cell-type-wilms-tumor-06/notebooks/03_clustering_exploration.Rmd
Debugs related to reorg into notebooks, and new data release which can have fewer than 500 cells

- `nbins` is the number of bins for `AddModuleScore` to use. Its default value is 24, which matches Seurat's default value.
```{r}
MSigDB_score <- function(object, category, gs_name, name, nbin = 24) {
  set <- msigdbr(species = "human", category = category)
  set_list <- set %>%
    dplyr::filter(gs_name == gs_name) %>%
    dplyr::distinct(gs_name, gene_symbol, human_ensembl_gene) %>%
    as.data.frame()
  set_list <- list(set_list$human_ensembl_gene)
  suppressWarnings({
    object <- AddModuleScore(object, features = set_list, name = name, nbin = nbin)
  })
  return(object)
}
```

@@

### Cell cycle information
```{r cell_cycle, eval=!params$testing}
s.genes <- srat@assays$RNA@meta.data$gene_ids[srat@assays$RNA@meta.data$gene_symbol %in% cc.genes$s.genes]
g2m.genes <- srat@assays$RNA@meta.data$gene_ids[srat@assays$RNA@meta.data$gene_symbol %in% cc.genes$g2m.genes]
srat <- CellCycleScoring(srat, s.features = s.genes, g2m.features = g2m.genes, set.ident = FALSE, nbin = seurat_nbins)
```

>>>>

- `nbins` is the number of bins for `AddModuleScore` to use. Its default value is 24, which matches Seurat's default value.
```{r}
MSigDB_score <- function(object, category, gs_name, name, nbin = 24) {
  set <- msigdbr(species = "human", category = category)
  set_list <- set %>%
    dplyr::filter(gs_name == gs_name) %>%
    dplyr::distinct(gs_name, gene_symbol, human_ensembl_gene) %>%
    as.data.frame()
  set_list <- list(set_list$human_ensembl_gene)
  suppressWarnings({
    # AddModuleScore is known to throw an error when there are too few cells, e.g.:
    # https://github.com/satijalab/seurat/issues/5814 (the source of this error is the call to AddModuleScore)
    # When we have fewer than ~500 cells, this error will occur. To avoid this, we override the `ctrl` (default 100) parameter to a lower value (15)
    if (ncol(object) <= 500) {
      object <- AddModuleScore(object, features = set_list, name = name, nbin = nbin, ctrl = 15)
    } else {
      object <- AddModuleScore(object, features = set_list, name = name, nbin = nbin)
    }
  })
  return(object)
}
```

@@

### Cell cycle information
```{r cell_cycle, eval=!params$testing}
s.genes <- srat@assays$RNA@meta.data$gene_ids[srat@assays$RNA@meta.data$gene_symbol %in% cc.genes$s.genes]
g2m.genes <- srat@assays$RNA@meta.data$gene_ids[srat@assays$RNA@meta.data$gene_symbol %in% cc.genes$g2m.genes]

# CellCycleSorting is known to throw an error when there are too few cells, e.g.:
# https://github.com/satijalab/seurat/issues/5814
# When we have fewer than ~500 cells, this error will occur. To avoid this, we override the `ctrl` (default 100) parameter to a lower value (15)
if (ncol(srat) <= 500) {
  srat <- CellCycleScoring(srat, s.features = s.genes, g2m.features = g2m.genes, set.ident = FALSE, nbin = seurat_nbins, ctrl = 15)
} else {
  srat <- CellCycleScoring(srat, s.features = s.genes, g2m.features = g2m.genes, set.ident = FALSE, nbin = seurat_nbins)
}
```

### 2024-11-26 analyses/cell-type-wilms-tumor-06/supplemental-notebooks/compare-label-transfer-approaches.Rmd
Debugs related to reorg into notebooks, and new data release which can have fewer than 500 cells

# functions to perform label transfer with azimuth-adapted approach
source(
  file.path(module_base, "notebook_template", "utils", "label-transfer-functions.R")
)

>>>>

# functions to perform label transfer with azimuth-adapted approach
source(
  file.path(module_base, "notebooks", "utils", "label-transfer-functions.R")
)
